Improving the portuguese hyphenation rules

### 2024-06-14 article/dictionary.tex
updates

For those hyphenations with four votes, we adopted a similar approach to the
one used previously. Many words garnered only four votes, indicating a lack of
alternative hyphenations provided by any dictionary, and were thus initially
hyphenated as proparoxytones. These instances were typically distinguished by
diacritics in the third-to-last syllable and a \emph{V-V} sequence at the end.
To address this, we converted these words into paroxytones by eliminating the
final hyphen. In cases where alternative hyphenations were available, we
assessed the presence of diacritics. If diacritics were detected, we chose the
paroxytone option; otherwise, we defaulted to the proparoxytone option. In
instances with multiple alternative options, we selected the option with the
highest frequency of votes. This procedure is implemented in the script
\texttt{get4.sh}.

>>>>

For those hyphenations with four votes, we adopted a similar approach to the
one used previously. Many words garnered only four votes, indicating a lack of
alternative hyphenations provided by any dictionary, and they were thus initially
hyphenated as proparoxytones. These instances were typically distinguished by
diacritics in the third-to-last syllable and a \emph{V-V} sequence at the end.
To address this, we converted these words into paroxytones by eliminating the
final hyphen. In cases where alternative hyphenations were available, we
assessed the presence of diacritics. If diacritics were detected, we chose the
paroxytone option; otherwise, we defaulted to the proparoxytone option. In
instances with multiple alternative options, we selected the option with the
highest frequency of votes. This procedure is implemented in the script
\texttt{get4.sh}, and the resulting list has \num{\NumberOfFourAgrees} words.

### 2024-06-14 article/limitations.tex
updates

\begin{description} 
    \item[Morphological determination:\label{morpho-det}] Prefixes such as \emph{re-},
	\emph{sub-}, \emph{ciber-}, \emph{hiper-}, and \emph{auto-}, among others, require separating the
	prefix from its stem, which can lead to phonological issues. For example, words like
	\emph{reiniciar}, \emph{sublinhar}, \emph{ciberespaço}, \emph{hiperalgesia}, and \emph{autoimagem} contain prefixes and could be hyphenated as 
	\emph{re-i-ni-ci-ar}, \emph{sub-li-nhar}, \emph{ci-ber-es-pa-ço}, \emph{hi-per-al-ge-si-a}, and \emph{au-to-i-ma-gem}, 
	respectively, to respect their morphological formation. However,
        considering that Portuguese hyphenates its words based on syllabic phonological correlates, 
	words that require morphological information, such as these, might not have their hyphenation performed correctly. 
	%In our corpus, there are twelve such words (eight with \emph{re-}, and one each with \emph{auto-}, \emph{meta-}, \emph{sub-}, \emph{cyber-}, \emph{geo-}, or \emph{neuro-}). 
	These examples are presented in \Cref{tab-mor-syl}.
	% $ { head -n 1 ../data/hyphenations.csv; grep "^sublinhar," ../data/hyphenations.csv; } | column -t -s,
	% su-bli-nhar (priberam, wiktionary, aulete)
	% sub-li-nhar (michaelis, portal, dicio)

@@

    \item[Foreignness:\label{foreignness}] There is a group of words %20 words 
	in the corpus that are terminologies or words incorporated into the 
	Portuguese language without full phonological adaptation, such as 
	\emph{darwinismo}, \emph{quilowatt}, and \emph{esfiha}. The lack of
	adaptation makes the phonological pattern very specific to the word,
	making it impossible to incorporate their cases \TeX{}'s rules alongside the
	other rules. The solution is to add them to the exception word list.

    \item[Word-initial consonant clusters:\label{word-init-cc}] Portuguese has few cases of
	consonant clusters at the beginning of a word. They are, in general, 
	etymological remnants and are currently unproductive in the language,
	since there are no neologisms with this pattern. Encounters like
	\emph{ps-} and \emph{pn-} are more frequent, as they are present in
	words like \emph{psicologia} and \emph{pneu}, which have moderate 
	frequency in the language. These can be predicted by specific rules that would cover  
	49 and 13 words in the corpus, respectively.
	However, there are consonant clusters that are
	found in very specific and low-frequency words. Although possible, it
	is not worth adding very specific rules for clusters found in words like
	\emph{dzeta}, \emph{gnu}, \emph{cnidário},
	\emph{ftálico}, and \emph{gnaisse} -- which amount to only five words.

    \item[Abbreviations, Acronyms, or Initialisms:\label{abbrev}] 
       Whether for efficiency, convenience, clarity, or specialized jargon, 
       it is common to use shortened versions of words or phrases. \emph{Abbreviation} is 
       a method employed to achieve this shortening. In our Portuguese corpus, we find examples such as 
       \emph{etc.}\footnote{Latin expression \emph{et cetera}, meaning `and other similar things'.}, 
       \emph{Dr.}\footnote{\emph{doutor} (doctor, person with PhD title, but popularly used to designate an erudite individual)}, 
       \emph{Exmo.}\footnote{\emph{Excelentíssimo} (honourable)},
       \emph{cap.}\footnote{\emph{capítulo} (chapter)}, \emph{Univ.}\footnote{Universidade (university)}, 
       \emph{ed.}\footnote{\emph{edição} (edition)}, 
       \emph{s.n.}\footnote{sine nomine, Latin expression meaning `without a name', mostly used in the context of publishing.}.
       Another shortened form is an \emph{initialism}, which consists of using
       the initial letters of words to create a shortened version. However,
       initialisms may not always conform to the hyphenation rules described in
       this work, as they do not necessarily follow the orthographic or
       phonotactic standards of the language. Some abbreviations found in the
       corpus include 
       \emph{SESC}\footnote{Serviço Social do Comércio}, 
       \emph{INSS}\footnote{\emph{Instituto Nacional do Seguro Social} (National Institute of Social Security)}, 
       \emph{PCdoB}\footnote{\emph{Partido Comunista do Brasil} (Communist Party of Brazil)},
       \emph{PM}\footnote{\emph{Polícia Militar} (military police)}, and 
       \emph{UFRJ}\footnote{Universidade Federal do Rio de Janeiro}.
       \emph{Acronyms} are a specific type of shortening where the first letters (or
       groups of letters) of each word are combined to form a new pronounceable
       word. In the corpus, we encounter examples like 
       \emph{Anatel}\footnote{\emph{Agência Nacional de Telecomunicações} (National Telecommunications Agency)},
       \emph{Ovni}\footnote{\emph{Objeto voador não identificado} (unidentified flying object - UFO)},
       \emph{Sida}\footnote{\emph{Síndrome da Imunodeficiência Adquirida} (acquired immunodeficiency syndrome - AIDS)} (in Portugal) and
       \emph{Mercosul}\footnote{\emph{Mercado Comum do Sul} (Southern Common Market).}.
       These various shortened forms play an important role in written
       language, providing concise ways to represent longer words or phrases.
       It is important to note that abbreviations, acronyms, and initialisms
       are generally treated as single units and are not hyphenated.

>>>>

\begin{description} 
    \item [Morphological determination\label{morpho-det}] Prefixes such as \emph{re-},
	\emph{sub-}, \emph{ciber-}, \emph{hiper-}, and \emph{auto-}, among others, require separating the
	prefix from its stem, which can lead to phonological issues. For example, words like
	\emph{reiniciar}, \emph{sublinhar}, \emph{ciberespaço}, \emph{hiperalgesia}, and \emph{autoimagem} contain prefixes and could be hyphenated as 
	\emph{re-i-ni-ci-ar}, \emph{sub-li-nhar}, \emph{ci-ber-es-pa-ço}, \emph{hi-per-al-ge-si-a}, and \emph{au-to-i-ma-gem}, 
	respectively, to respect their morphological formation. However,
        considering that Portuguese hyphenates its words based on syllabic phonological correlates, 
	words that require morphological information, such as these, might not have their hyphenation performed correctly. 
	%In our corpus, there are twelve such words (eight with \emph{re-}, and one each with \emph{auto-}, \emph{meta-}, \emph{sub-}, \emph{cyber-}, \emph{geo-}, or \emph{neuro-}). 
	These examples are presented in \Cref{tab-mor-syl}.
	% $ { head -n 1 ../data/hyphenations.csv; grep "^sublinhar," ../data/hyphenations.csv; } | column -t -s,
	% su-bli-nhar (priberam, wiktionary, aulete)
	% sub-li-nhar (michaelis, portal, dicio)

@@

    \item [Foreignness\label{foreignness}] There is a group of words %20 words 
	in the corpus that are terminologies or words incorporated into the 
	Portuguese language without full phonological adaptation, such as 
	\emph{darwinismo}, \emph{quilowatt}, and \emph{esfiha}. The lack of
	adaptation makes the phonological pattern very specific to the word,
	making it impossible to incorporate their cases \TeX{}'s rules alongside the
	other rules. The solution is to add them to the exception word list.

    \item [Word-initial consonant clusters\label{word-init-cc}] Portuguese has few cases of
	consonant clusters at the beginning of a word. They are, in general, 
	etymological remnants and are currently unproductive in the language,
	since there are no neologisms with this pattern. Encounters like
	\emph{ps-} and \emph{pn-} are more frequent, as they are present in
	words like \emph{psicologia} and \emph{pneu}, which have moderate 
	frequency in the language. These can be predicted by specific rules that would cover  
	49 and 13 words in the corpus, respectively.
	However, there are consonant clusters that are
	found in very specific and low-frequency words. Although possible, it
	is not worth adding very specific rules for clusters found in words like
	\emph{dzeta}, \emph{gnu}, \emph{cnidário},
	\emph{ftálico}, and \emph{gnaisse} -- which amount to only five words.

    \item [Abbreviations, Acronyms, or Initialisms\label{abbrev}] 
       Whether for efficiency, convenience, clarity, or specialized jargon, 
       it is common to use shortened versions of words or phrases. \emph{Abbreviation} is 
       a method employed to achieve this shortening. In our Portuguese corpus, we find examples such as 
       \emph{etc.}\footnote{Latin expression \emph{et cetera}, meaning `and other similar things'.}, 
       \emph{Dr.}\footnote{\emph{doutor} (doctor, person with PhD title, but popularly used to designate an erudite individual)}, 
       \emph{Exmo.}\footnote{\emph{Excelentíssimo} (honourable)},
       \emph{cap.}\footnote{\emph{capítulo} (chapter)}, \emph{Univ.}\footnote{Universidade (university)}, 
       \emph{ed.}\footnote{\emph{edição} (edition)}, 
       \emph{s.n.}\footnote{sine nomine, Latin expression meaning `without a name', mostly used in the context of publishing.}.
       Another shortened form is an \emph{initialism}, which consists of using
       the initial letters of words to create a shortened version. However,
       initialisms may not always conform to the hyphenation rules described in
       this work, as they do not necessarily follow the orthographic or
       phonotactic standards of the language. Some abbreviations found in the
       corpus include 
       \emph{SESC}\footnote{Serviço Social do Comércio}, 
       \emph{INSS}\footnote{\emph{Instituto Nacional do Seguro Social} (National Institute of Social Security)}, 
       \emph{PCdoB}\footnote{\emph{Partido Comunista do Brasil} (Communist Party of Brazil)},
       \emph{PM}\footnote{\emph{Polícia Militar} (military police)}, and 
       \emph{UFRJ}\footnote{Universidade Federal do Rio de Janeiro}.
       \emph{Acronyms} are a specific type of shortening where the first letters (or
       groups of letters) of each word are combined to form a new pronounceable
       word. In the corpus, we encounter examples like 
       \emph{Anatel}\footnote{\emph{Agência Nacional de Telecomunicações} (National Telecommunications Agency)},
       \emph{Ovni}\footnote{\emph{Objeto voador não identificado} (unidentified flying object - UFO)},
       \emph{Sida}\footnote{\emph{Síndrome da Imunodeficiência Adquirida} (acquired immunodeficiency syndrome - AIDS)} (in Portugal) and
       \emph{Mercosul}\footnote{\emph{Mercado Comum do Sul} (Southern Common Market).}.
       These various shortened forms play an important role in written
       language, providing concise ways to represent longer words or phrases.
       It is important to note that abbreviations, acronyms, and initialisms
       are generally treated as single units and are not hyphenated.

### 2024-06-14 article/patgenrules.tex
updates

The parameters mentioned in \Cref{sec-patgen} to run \emph{patgen} are provided interactively during the execution of \emph{patgen}.
We might automatize this process using a script. The \Cref{lst-patgentest} presents the code block
responsible for passing these parameters to \emph{patgen}.
\lstinputlisting[language={bash}, linerange={76-83}, basicstyle=\small\ttfamily, breaklines=true, caption={Portuguese translation file.}, label={lst-patgentest}]{../scripts/patgentest.sh}
As an example of execution, when choosing \texttt{hyph\_start}=1, \texttt{hyph\_finish}=9, \texttt{pat\_start}=2, \texttt{pat\_finish}=5,
\texttt{good weight}=1, \texttt{bad weight}=1, and \texttt{threshold}=1, we achieve 99.97\%(86799) correct 
hyphenations, 0.01\%(12) incorrect hyphenations, and 0.03\%(25) missing hyphenations.
It produced a set of 799 rules, where just 123 are present in the handcrafted set of rules.

We have also used the approach used by \textcite{sojka2019} in creating the script \texttt{make-full-pattern}.
In this approach, the parameters are loaded from a file and might be distinct for each level.
We start from an empty set and create progressive higher order rules, always using the rules from the previous
step as the starting point at the next step.
Using this approach with the 8 level parameters they used to create hyphenation rules to German,
we achieved 99.98\%(86820) correct hyphenations, 0.01\%(9) incorrect hyphenations, and 0.00\%(4) missed hyphenations.
The result was a set of 593 rules, with only 69 in common with those handcrafted rules proposed earlier.

@@

The results from the set of rules crafted by \emph{patgen} seems outstanding, but we must be 
skeptical of its generalization power. To inquire on this matter, we decided to retrieve the
list of words where just three hyphenations were found in the dictionaries. We have also dealt
with apparent proparoxytones, to be consistent in our approach.  
In this new set, the number of correct, incorrect and missing hyphenations for each set of rules
is given \Cref{tbl-comres-hy3}.
\begin{table}[h]
\centering
\caption{Compared results between two use cases of \emph{patgen} and the handcrafted rule set.}\label{tbl-comres-hy3}
\begin{tabular}{llll}
rule set          & correct & incorrect & missing \\
\hline
fixed parameters  & 7465    & 150       & 183 \\
make-full-pattern & 647     & 47        & 7089 \\
handcrafted       & 7335    & 42        & 385 
\end{tabular}
\end{table}
It is important to hightlight that, among those 385 missing hyphenations in the handcrafted rules,
278 cases correspond to the last hyphen, leaving a single vowel as a final syllable.
Comparing to the other set of rules, we have 7 (in 183, for fixed parameters case), and 257 (in 7089, 
for the \texttt{make-full-pattern} script).

>>>>

The parameters mentioned in \Cref{sec-patgen} to run \emph{patgen} are provided interactively during the execution of \emph{patgen}.
We might automatize this process using a script. The \Cref{lst-patgentest} presents the code block
responsible for passing these parameters to \emph{patgen}.
\lstinputlisting[language={bash}, linerange={76-83}, basicstyle=\small\ttfamily, breaklines=true, caption={Portuguese translation file.}, label={lst-patgentest}]{../scripts/patgentest.sh}
As an example of execution, when choosing \texttt{hyph\_start}=1, \texttt{hyph\_finish}=9, \texttt{pat\_start}=2, \texttt{pat\_finish}=5,
\texttt{good weight}=1, \texttt{bad weight}=1, and \texttt{threshold}=1, we achieve 99.97\% (86799) correct 
hyphenations, 0.01\% (12) incorrect hyphenations, and 0.03\% (25) missing hyphenations.
It produced a set of 799 rules, where just 123 are present in our handcrafted set of rules.

We have also used the approach used by \textcite{sojka2019} in creating the script \texttt{make-full-pattern}.
In this approach, the parameters are loaded from a file and might be distinct for each level.
We start from an empty set and create progressive higher order rules, always using the rules from the previous
step as the starting point at the next step.
Using this approach with the 8 level parameters they used to create hyphenation rules to German,
we achieved 99.98\% (86820) correct hyphenations, 0.01\% (9) incorrect hyphenations, and 0.00\% (4) missed hyphenations.
The result was a set of 593 rules, with only 69 in common with those handcrafted rules proposed earlier.

@@

The results from the set of rules crafted by \emph{patgen} seems outstanding, but we must be 
skeptical of its generalization power. To inquire on this matter, we decided to retrieve the
list of words where just three hyphenations were found in the dictionaries. We have also dealt
with apparent proparoxytones, to be consistent in our approach.  
In this new set, the number of correct, incorrect and missing hyphenations for each set of rules
is given \Cref{tbl-comres-hy3}.

\begin{table}[h]
\centering
\caption{Compared results between two use cases of \emph{patgen} and the handcrafted rule set when hyphenating the set of words from the dictionary built from those words with 3 agreements on the online dictionaries.}\label{tbl-comres-hy3}
\begin{tabular}{llll}
rule set          & correct & incorrect & missing \\
\hline
fixed parameters  & 7465    & 150       & 183 \\
make-full-pattern & 647     & 47        & 7089 \\
handcrafted       & 7335    & 42        & 385 
\end{tabular}
\end{table}

It is important to hightlight that, among those 385 missing hyphenations in the handcrafted rules,
278 cases correspond to the last hyphen, leaving a single vowel as a final syllable.
Comparing to the other set of rules, we have 7 (in 183, for fixed parameters case), and 257 (in 7089, 
for the \texttt{make-full-pattern} script).

### 2024-06-14 article/resultsdefaultrules.tex
updates

In this section we will apprise the performance of default \TeX{} hyphenation
rules on the list of hyphenated words, which creation was described in
\Cref{sec-dictionay}. \Cref{tab-default-results} summaries the results, showing
the number of words correctly hyphenated words by the default rules, the number
of words wrongly hyphenated and the number of words where a hyphenation point
was missed.

\begin{table}[htbp]
\centering
\caption{Results of \TeX{} default hyphenation rules on \texttt{hyphenations6}, \texttt{hyphenations5} and \texttt{hyphenations4} dictionaries.}
\label{tab-default-results}
\begin{tabular}{lrrrr}
\toprule
word list 	& \# correct 		& \# wrong 		& \# missing 		& \# entries 		\\
\midrule
hyphenations6 	& \NumberOfCorrectSix{}	& \NumberOfWrongSix{}	& \NumberOfMissingSix{}	& \NumberOfTotalSix{} 	\\
hyphenations5 	& \NumberOfCorrectFive{}& \NumberOfWrongFive{}	& \NumberOfMissingFive{}& \NumberOfTotalFive{} 	\\
hyphenations4 	& \NumberOfCorrectFour{}& \NumberOfWrongFour{}	& \NumberOfMissingFour{}& \NumberOfTotalFour{} 	\\
total		& \NumberOfCorrectSum{} & \NumberOfWrongSum{} 	& \NumberOfMissingSum{}	& \NumberOfTotalSum{}   \\
\bottomrule
\end{tabular}
\end{table}

For the list comprised in \texttt{hyphenations6}, below is the complete list of
\NumberOfWrongSix{} words that were incorrectly hyphenated:
% wrong
\begin{multicols}{3}
\setlength{\columnsep}{0pt}
\setlength{\parindent}{0pt}
\emph{
    \input{/tmp/hyphenations6result-default-patterns_wrong.dic}
}
\end{multicols}
In each instance, an erroneous hyphenation point was placed at the beginning of
a word, indicating a potential issue in recognizing certain prefixing morphemes
in the language.  Among those, 20 occurred in words containing the \emph{psi}
morpheme, where the algorithm erroneously hyphenated between the initial
\emph{p} and the following \emph{s}. Additionally, 3 cases involved the
starting \emph{gno} sequence, while the remaining errors were observed in the
sequences \emph{pneu} (3), \emph{tch} (1), \emph{tme} (1), \emph{pto} (1), and
\emph{sub} (1).

Those erroneous hyphenation points might be corrected by the introduction of a few rules:
\texttt{.p2si, .p2sí} (see \cref{rulegrp_psi}), \texttt{.g2no, .g2nó} (see \cref{rulegrp_gno}), 
\texttt{p2neu} (see \cref{rulegrp_pneu}),
\texttt{t2c} (see \cref{rulegrp_tc}), \texttt{.t2m} (see \cref{rulegrp_tm}), \texttt{.p2t} (see \cref{rulegrp_pt}) 
and \texttt{su2b3r}, \texttt{su2b3l}  (see \cref{rulegrp_sub}). They are deeply 
discussed in \Cref{sec-hyph-rpatches}.

% $ ../scripts/resultfileexamples.sh -m -f ../data/hyphenations6result-default-patterns.dic | grep -o "[^-]-[^-]" | sort | uniq -c | sort -nr | column

The first 20 examples of missing hyphenations are displayed in the following list:
% missing
\begin{multicols}{3}
\setlength{\columnsep}{0pt}
\setlength{\parindent}{0pt}
\emph{
\input{/tmp/hyphenations6result-default-patterns_missing.dic}
...
}
\end{multicols}
It is a long list with \NumberOfMissingSix{} entries, therefore it is
unavoidable to analyse it through clusters of certain patterns. It is presented
below the counts of immediate context in which those missing hyphenations were
found:
\begin{multicols}{5}
\setlength{\columnsep}{0pt}
\setlength{\parindent}{0pt}
\emph{
\input{/tmp/hyphenations6result-default-patterns_missing_context}
}
\end{multicols}

% number of words correctly hyphenated using default TeX rules when analysing the words in hyphenations6.dic
%\captureshell*[\NumberOfCorrectSix]{../scripts/resultfilemetrics.sh -f ../data/hyphenations6result-default-patterns.dic -c}

% number of words with wrong hyphenations using default TeX rules when analysing the words in hyphenations6.dic
%\captureshell*[\NumberOfWrongSix]{../scripts/resultfilemetrics.sh -f ../data/hyphenations6result-default-patterns.dic -w}

% number of words with missing hyphenations using default TeX rules when analysing the words in hyphenations6.dic
%\captureshell*[\NumberOfMissingSix]{../scripts/resultfilemetrics.sh -f ../data/hyphenations6result-default-patterns.dic -m}

% tra-b.a*lho
%    ^ ^ ^
%    | | |
%    | | + -- correct (*)
%    | + ---- wrong   (.)  
%    + ------ missed  (-)

On the top of the list we see the missing hyphen in \emph{u-i}, accounting for
38 cases. Among those, 34 were originated from the pattern \texttt{u-ir} in the
end of word and 13 from \texttt{a-ir} in the end of word. 
To fix them, We will include the rules \texttt{u1ir.} and \texttt{a1ir.} (see 
\cref{rulegrp_air}). The cases that are not covered by this rule are:
\emph{tu-im}, \emph{je*su-i*tis*mo}, \emph{ma*lau-i*a*no}, and
\emph{con*tri*bu-in*te}. On \Cref{sec-hyph-rpatches} we see how to deal with
them.

The default rules already include 17 separating rules for vowel sequences:
% vowel="[aeiouáãàâéêíóõôú]"; grep "$vowel[13579]$vowel" ../data/default.TeX.pt-br.patterns | sed ':a;N;$!ba;s/\n/ \\\\ /g'
\begin{multicols}{5}
\setlength{\columnsep}{0pt}
\setlength{\parindent}{0pt}
\texttt{a3a \\ a3e \\ a3o \\ e3a \\ e3e \\ e3o \\ i3a \\ i3e \\ i3i \\ i3o \\ i3â \\ i3ê \\ i3ô \\ o3a \\ o3e \\ o3o \\ u3a \\ u3e \\ u3o \\ u3u}
\end{multicols}
but there are just 3 rules to hyphenate between vowels when one has a diacritic: \texttt{i3â}, \texttt{i3ê}, \texttt{i3ô}.
We could then add more 29 rules to account for the missing hyphenations between vowels with diacritics and also the
missing rule \texttt{i1u}:
% ../scripts/resultfileexamples.sh -m -f ../data/hyphenations6result-default-patterns.dic | ../scripts/filtersurrounding.sh -c '-' -n 1 | sort | uniq -c | sort -nr | grep "[áàâãéêíóôõú]" | awk '{print $2}' | sort | sed ':a;N;$!ba;s/\n/ \\\\ /g'
\begin{multicols}{5}
\setlength{\columnsep}{0pt}
\setlength{\parindent}{0pt}
\texttt{
% complete set of 37 rules
%a1â \\ a1ã \\ a1é \\ a1í \\ a1ó \\ a1ô \\ a1ú \\ 
%e1á \\ e1â \\ e1ã \\ e1é \\ e1ê \\ e1í \\ e1ó \\ e1ô \\ e1ú \\ é1o \\ 
%i1á \\ i1ã \\ i1ã \\ i1é \\ i1í \\ i1ó \\ i1u \\ i1ú \\ í1a \\ 
%o1á \\ o1ã \\ o1é \\ o1ê \\ o1í \\ o1ó \\ 
%u1á \\ u1ã \\ u1é \\ u1ê \\ u1í \\ ú1o
%
% 29 rules necessary for hyphenations6 dic
a1é \\ a1í \\ a1ó \\ a1ú \\ 
e1á \\ e1â \\ e1ã \\ e1é \\ e1ê \\ e1í \\ e1ó \\ é1o \\ e1ú \\ 
i1á \\ i1ã \\ í1a \\ i1é \\ i1í \\ i1ó \\ í1o \\ i1u \\ i1ú \\ 
o1á \\ o1é \\ o1í \\ o1ó \\ 
u1á \\ u1ã \\ u1í \\ ú1o
}
%\emph{a-é \\ a-í \\ a-ó \\ a-ú \\ e-á \\ e-â \\ e-ã \\ e-é \\ e-ê \\ e-í \\ e-ó \\ é-o \\ e-ú \\ i-á \\ i-ã \\ í-a \\ i-é \\ i-í \\ i-ó \\ í-o \\ i-ú \\ o-á \\ o-é \\ o-í \\ o-ó \\ u-á \\ u-ã \\ u-í \\ ú-o}
\end{multicols}
See more about this on \cref{rulegrp_aa}. These rules will require exceptions for sequences with \emph{[qg]uV'} 
(where we used \emph{V'} to represent a vowel with a diacritic). See more on \cref{rulegrp_aa,rulegrp_gua}.

From the list of missing hyphenations above, we find another recurring pattern, a missing hyphen preceding \emph{q}: 
% ../scripts/resultfileexamples.sh -m -f ../data/hyphenations6result-default-patterns.dic | ../scripts/filtersurrounding.sh -c '-' -n 1 | sort | uniq -c | sort -nr | awk '{print $2}' | sort | grep "q$" | sort | sed ':a;N;$!ba;s/\n/ \\\\ /g'
\begin{multicols}{5}
\setlength{\columnsep}{0pt}
\setlength{\parindent}{0pt}
\emph{a-q \\ e-q \\ i-q \\ o-q \\ r-q \\ s-q \\ u-q}
\end{multicols}
That issue might be easily solved by adding a hyphenation rule: \texttt{1qu}, but as the \emph{q} is always followed by 
a sequence \emph{uV}, and there are already four hyphenation rules involving this sort of sequence in the default rules
(\texttt{1qu4a}, \texttt{1qu4e}, \texttt{1qu4i} and \texttt{1qu4o}), the rules 
% $ grep "q" ../data/patch.TeX.pt-br.patterns | awk '{print "\\texttt{" $1 "}"}' | sed ':a;N;$!ba;s/\n/, /g'
\texttt{1qu2á}, \texttt{1qu2é}, \texttt{1qu2ê}, \texttt{1qu2í}, \texttt{que2i}, \texttt{1qu4ã}, \texttt{1qu4ó}, \texttt{1qu4â}
could be added (see \cref{rulegrp_aa,rulegrp_gua,rulegrp_quo}) but they have not proven to be necessary.
%%% XXX check if they are not really necessary

The missing hyphens in \emph{a-v} and \emph{i-v} might be easily solved by introducing the rule \texttt{1vô} (see \cref{rulegrp_bo}) which 
complements the rule \texttt{1vo} already included in the default set of rules.
% pi-vô, a-vô, bi*sa-vô ...
Similarly, the missing hyphens \emph{e-l}, \emph{u-l}, \emph{o-l}, and \emph{a-l} are corrected by the introduction of the rule \texttt{1lô} rule (see \cref{rulegrp_bo}),
complementing the already included rule \texttt{1lo}.
In the same way, the rules \texttt{1cô}, \texttt{1gô}, \texttt{1bô}, \texttt{1tô}, \texttt{1rô}, and \texttt{1pô} (see \cref{rulegrp_bo}) 
should be added to fix the missing hyphens \emph{e-c}, \emph{i-c}, \emph{o-g}, \emph{o-b}, \emph{a-t}, \emph{a-r},
and \emph{a-p}, respectively.
% ca*me-lô, gi*go-lô, ..., bi*be-lô, su.b-lu*nar
%                                       rule 1b2l causes wrong hyphen
%
%    .   s   u   b   l   u   n   a   r   .
%     1   0   0   |   |   |   |   |   |    1su
%     |   |   1   2   0   |   |   |   |    1b2l
%     |   |   |   1   0   0   |   |   |    1lu
%     |   |   |   |   |   1   0   0   |    1na
%max: 1   0   1   2   0   1   0   0   0
%final: s   u - b   l   u - n   a   r
The missing hyphen \emph{b-l} happened in \emph{su.b-lu*nar}, which also has a wrong hyphenation.
Those errors were caused by the default rule \texttt{1b2l}. The introduction of rule \texttt{su2b3l} 
will help solve this matter (see details in \cref{rulegrp_sub}).

% e-c, i-c: missing rule 1cô
% o-g: missing rule 1gô
% o-b: missing rule 1bô
% a-t: missing rule 1tô
% a-r: missing rule 1rô
% a-p: missing rule 1pô

The remaining missing hyphen is \emph{o-w}, which comes from the foreign \emph{kilowatt}.
Those cases of words borrowed from other languages are dealt in the \ref{foreignness} topic ahead.
% o-w: qui*lo-watt

% number of words with missing hyphenations using default TeX rules when analysing the words in hyphenations6.dic

>>>>

In this section we will apprise the performance of default \TeX{} hyphenation
rules on the list of hyphenated words, which creation was described in
\Cref{sec-dictionay}. \Cref{tab-default-results} summaries the results, showing
the number of words correctly hyphenated words by the default rules, the number
of words incorrectly hyphenated and the number of words where a hyphenation point
was missed.

\begin{table}[htbp]
\centering
\caption{Results of \TeX{} default hyphenation rules on \texttt{hyphenations6}, \texttt{hyphenations5} and \texttt{hyphenations4} dictionaries.}
\label{tab-default-results}
\begin{tabular}{lrrrr}
\toprule
word list 	& \# correct 		& \# incorrect		& \# missing 		& \# entries 		\\
\midrule
hyphenations6 	& \NumberOfCorrectSix{}	& \NumberOfWrongSix{}	& \NumberOfMissingSix{}	& \NumberOfTotalSix{} 	\\
hyphenations5 	& \NumberOfCorrectFive{}& \NumberOfWrongFive{}	& \NumberOfMissingFive{}& \NumberOfTotalFive{} 	\\
hyphenations4 	& \NumberOfCorrectFour{}& \NumberOfWrongFour{}	& \NumberOfMissingFour{}& \NumberOfTotalFour{} 	\\
total		& \NumberOfCorrectSum{} & \NumberOfWrongSum{} 	& \NumberOfMissingSum{}	& \NumberOfTotalSum{}   \\
\bottomrule
\end{tabular}
\end{table}

Each word in the list is hyphenated using the set of default rules using the \texttt{gohyphen} script
(the \texttt{gohyphenfull} script is used to see in the interplay among all rules that apply to a given word) \parencite{gundlach2021}.
The hyphenation resulted from the given rules is compared to the hyphenation in our dictionary, so
we may check if there were erros (incorrect or missing hyphenation points).
